Runtime Feedback in a Meta-Tracing JIT for Efficient Dynamic Languages

### 2011-03-28 talk/icooolps2011/paper.tex
expand related work some more

To be able to do this recording, VMs with a
tracing JIT typically contain an interpreter. After a user program is
started the interpreter is used; only the most frequently executed paths through the user
program are turned into machine code. The tracing JIT tries to produce traces
that correspond to loops in the traced program, but most tracing JITs now also
have support for tracing non-loops \cite{XXX}.

@@

Tracing through the execution of an interpreter has many advantages. It makes
the tracer, its optimizers and backends reusable for a variety of languages. The
language semantics do not need to be encoded into the JIT. Instead the tracer
just picks them up from the interpreter. XXX mention disadvantage of long
traces?

@@

Partial evaluation \cite{jones_partial_1993} tries to automatically transform
interpreters into compilers using the second futamura projection
\cite{futamura_partial_1999}. XXX

@@

Somewhat relatedly, the proposed ``invokedynamic'' bytecode
\cite{rose_bytecodes_2009} that will be added to the JVM is supposed to make the
implementation of dynamic languages on top of JVMs easier. The bytecode gives
the language implementor control over how the JIT optimizes the language's
features. XXX

%We already explored promotion in other context, such as earlier examples of
%promotion \cite{carl_friedrich_bolz_towards_????}, \cite{armin_rigo_jit_2007}

\cite{chambers_efficient_1989} maps
\cite{hoelzle_optimizing_1994} Type feedback
\cite{hoelzle_optimizing_1991} PICs
find the web page of V8 about maps

@@

jruby used versions at some point, jvm-l mailing list discusses them

>>>>

To be able to do this recording, VMs with a
tracing JIT typically contain an interpreter. After a user program is
started the interpreter is used; only the most frequently executed paths through the user
program are turned into machine code. The tracing JIT tries to produce traces
that correspond to loops in the traced program, but most tracing JITs now also
have support for tracing non-loops \cite{andreas_gal_incremental_2006}.

@@

Tracing through the execution of an interpreter has many advantages. It makes
the tracer, its optimizers and backends reusable for a variety of languages. The
language semantics do not need to be encoded into the JIT. Instead the tracer
just picks them up from the interpreter.

@@

Partial evaluation \cite{jones_partial_1993} tries to automatically transform
interpreters into compilers using the second futamura projection
\cite{futamura_partial_1999}. Given that classical partial evaluation works
strictly ahead of time, it inherently cannot support runtime feedback.

@@

Somewhat relatedly, the proposed ``invokedynamic'' bytecode
\cite{rose_bytecodes_2009} that will be added to the JVM is supposed to make the
implementation of dynamic languages on top of JVMs easier. The bytecode gives
the language implementor control over how the JIT optimizes the language's
features and when optimized code needs to be deoptimized. XXX

We already explored promotion in other context, such as earlier versions of
PyPy's JIT \cite{armin_rigo_jit_2007} as well as a Prolog partial evaluator
\cite{carl_friedrich_bolz_towards_????}. Promotion is quite similar to
(polymorphic) inline caching and runtime type feedback techniques which were
first used in Smalltalk \cite{deutsch_efficient_1984} and SELF
\cite{hoelzle_optimizing_1991,hoelzle_optimizing_1994} implementations.
Promotion is more general because any information can be cached in line, not
just classes of method receivers.

@@

%jruby used versions at some point, jvm-l mailing list discusses them
